Pollution, agricultural productivity, and development: Evidence from coal in plants in India\footnote{All scripts used in this paper are publicly available on my GitHub page: github.com/JoshMerfeld/pollution\_development}

### 2023-07-20 draft/draft.Rmd
updating draft

---
title: "Air pollution and agricultural productivity in a developing country^[I thank Ariel Ortiz-Bobea, Marc Bellemare, and David Sungho Park for comments and suggestions that greatly improved the quality of the paper, as well as conference attendees at the 2023 KER International Conference. The usual disclaimer applies. All scripts used in this paper are publicly available on my GitHub page: github.com/JoshMerfeld/pollution_development]"
author:
  - Joshua D. Merfeld^[KDI School of Public Policy and Management and IZA; merfeld@kdis.ac.kr]
date: "`r Sys.Date()`"
abstract: |
  | \noindent I document negative externalities of air pollution in the Indian agricultural sector. Using variation in pollution induced by changes in wind across years, I show that higher levels of pollution lead to decreased agricultural productivity, with large changes in productivity being common. The negative effects of pollution are larger in areas growing more labor-intensive crops, indicating that the pollution works at least partly through direct effects on labor productivity. Finally, combining wind direction with the rollout of coal plants, results indicate that pollution from coal plants has a larger effect on agricultural productivity than other types of pollution. Given that the agricultural sector is a refuge for the poor in many developing countries, these results suggest that the negative externalities of pollution may hit the poorest particularly hard.
  |
  | \textbf{\textit{Keywords}}: pollution, productivity, agriculture, labor, India
  | \textbf{\textit{JEL Codes}}: H40, I15, J22, O13, Q52, Q53

@@

@@

While reliance on coal is declining in much of the world, it continues to be an important source of energy in many countries. In 2021, energy generation from coal reached an all-time high,[^1] driven in part by rapid population growth in developing countries, including India. From 1990 to 2010, India's population grew by more than 40 percent, leading to a large increase in demand for power.[^2] While natural gas has become a popular alternative in much of the world, the Indian government met much of the increased demand through the construction of coal power plants. During those same two decades, power generation from all coal units of at least 30 megawatts in the country more than doubled, from 42.4 gigawatts to more than 100 gigawatts.[^3]

@@

While there may be positive effects from this increase in power generation -- for example, electricity may lead to economic and productivity growth [@dinkelman2011effects; @kline2014local; @rud2012electricity; @van2017long], though electrification alone may not be a sufficient condition [@lee2020does] -- these improvements do not come without downsides, especially when it comes to power derived from coal. Emissions from the production and burning of coal are also directly harmful to human health and the environment; pollutants like sulfur dioxide, nitrogen oxides, methane, and mercury are particularly harmful. While the use of coal had been declining, recent events -- including Russia's invasion of Ukraine -- have led to an increase in the demand for coal.[^5]

@@

In addition to pollution from coal, pollution more generally is also increasing in much of the world. Carbon dioxide equivalent per capita has almost double over the last 50 years in India, more than doubled in Indonesia, and almost tripled in China.^[https://edgar.jrc.ec.europa.eu] Much of this increase is due to the development process itself -- as countries develop, their economies transform, with more people moving out of agriculture and into manufacturing and services [@herrendorf2014growth]. Despite this development, agriculture remains a key sector in many developing countries, employing a large proportion of the poor [@dercon2009rural]. This also means agriculture is a key sector when it comes to poverty reduction [@christiaensen2011evolving]. Given the large effects of pollution on health and previously documented impacts on agriculture, this raises a important question: do rising pollution levels in developing countries threaten agricultural productivity?

Although we have made much progress over the last several decades, poverty and inequality remain important challenges in the developing world [@alvaredo2015recent]. Poverty is particularly prevalent in rural areas, where the majority of the world's poor live [@dercon2009rural], and in the agricultural sector, where the majority of the world's poor work [@christiaensen2011evolving]. As such, agricultural productivity is a key determinant of poverty in developing countries [@dercon2009rural], meaning that any threat to agricultural productivity is also a threat to poverty reduction.

@@

Using daily wind data and these spatial variables, I create an identifier separately for every village for whether a given village is located downwind from a coal plant on any given day. I then aggregate this exposure variable based on the desired analysis; for agricultural productivity, for example, I create an exposure variable measuring the total number of exposure days in the five-month agricultural season, focusing only on the monsoon (Kharif) season. More days being located downwind from a high-pollution location leads to substantially higher average concentrations of particulate matter 2.5 (PM 2.5).^[PM 2.5 is defined as particulate matter with a diameter of less than 2.5 micrometers.]

@@

The final set of analyses tries to disentangle the effects of pollution from the effects of coal pollution, specifically, by using wind direction and the rollout of coal plants across the country. Combining instrumental variables with a differences-in-differences strategy, I show that agricultural productivity has an elasticity of around -0.22 with respect to non-coal pollution. The opening of a coal plant, however, doubles the elasticity, to more than -0.44, indicating that coal plant emissions are particularly harmful to agricultural productivity, at least in India. This is consistent with coal emissions being quite different from other forms of pollution; coal emissions contain different types and quantities of harmful substances, meaning effects may differ greatly.

@@

This paper contributes to several strands of literature. We already have evidence of effects of pollution on different forms of labor activity. For example, general levels of pollution decrease productivity of call center works in China [@chang2019effect], lead people to perform worse on cognitive functions [@ebenstein2016long; @wen2022lower], lower labor market earnings [@borgschulte2022air], and also lead to lower levels of farm labor productivity in California, driven by changes in ozone concentration [@graff2012impact]. Also in China, @he2019severe find small but significant negative effects of pollution on manufacturing productivity, while @chen2022effect show pollution leads to labor out migration. Relevant to my results, @hanna2015effect show that decreases in pollution lead to increases in labor allocation in another developing country.

>>>>

---
title: "Air pollution and agricultural productivity in a developing country^[I thank Ariel Ortiz-Bobea, Marc Bellemare, and David Sungho Park for comments and suggestions that greatly improved the quality of the paper, as well as conference attendees at the 2023 KER International Conference and the University of Reading. The usual disclaimer applies. All scripts used in this paper are publicly available on my GitHub page: github.com/JoshMerfeld/pollution_development]"
author:
  - Joshua D. Merfeld^[KDI School of Public Policy and Management and IZA; merfeld@kdis.ac.kr]
date: "`r Sys.Date()`"
abstract: |
  | \noindent I document negative externalities of air pollution in the Indian agricultural sector. Using variation in pollution induced by changes in wind across years, I show that higher levels of pollution lead to decreased agricultural productivity. The negative effects of pollution are larger in areas growing more labor-intensive crops, indicating that the pollution works partly through direct effects on labor productivity. I also show that pollution from coal plants has a larger effect on agricultural productivity than other types of pollution. These results suggest that the negative externalities of pollution may hit the poorest particularly hard, given their reliance on agriculture.
  |
  | \textbf{\textit{Keywords}}: pollution, productivity, agriculture, labor, India
  | \textbf{\textit{JEL Codes}}: H40, I15, J22, O13, Q52, Q53

@@

<!-- 
Jonathan's comments:
- Drop "in a developing country" from the title?
- Add a sense of magnitudes, specifically the effects of pollution on yields. Like "pollution has increased on avearge by X percent in India, and this has led to a Y percent decrease in yields"
  - This is important
- Build up labor productivity bit a little more
  - Jonathan thinks current writing makes it sound more like a technical point than a big deal (which it is)
- Look at pollution prior to season? Kind of a "placebo" test
- Maybe move labor and coal parts up in the introduction
- Add a citation on the difference between PM 2.5 and PM 10
  - Jonathan says he didn't understand the paragraph on PM 2.5 until we discussed it. Important to make sure the reader understands it.
-->

@@

In 2021, energy generation from coal reached an all-time high,[^1] driven in part by rapid population growth in developing countries, including India. While natural gas has become a popular alternative in much of the world, the Indian government met much of the increased demand through the construction of coal power plants. From 1990 to 2010, India's population grew by more than 40 percent, leading to a large increase in demand for power.[^2] During those same two decades, power generation from all coal units of at least 30 megawatts in the country more than doubled, from 42.4 gigawatts to more than 100 gigawatts.[^3]

@@

While electricity may lead to economic and productivity growth [@dinkelman2011effects; @kline2014local; @rud2012electricity; @van2017long], though electrification alone may not be a sufficient condition [@lee2020does], these improvements do not come without downsides, especially when it comes to power derived from coal. Emissions from the production and burning of coal -- pollutants like sulfur dioxide, nitrogen oxides, methane, and mercury, for example -- are also directly harmful to human health and the environment. Although the use of coal had been declining, recent events -- including Russia's invasion of Ukraine -- have led to an increase in the demand for coal.[^5]

@@

Pollution more generally is also increasing in much of the world. Carbon dioxide equivalent per capita has almost doubled over the last 50 years in India, more than doubled in Indonesia, and almost tripled in China.^[https://edgar.jrc.ec.europa.eu] Much of this increase is due to the development process itself -- as countries develop, their economies transform, with more people moving out of agriculture and into manufacturing and services [@herrendorf2014growth]. Despite this development, agriculture remains a key sector in many developing countries, employing a large proportion of the poor [@dercon2009rural], and agriculture is a key sector for poverty reduction [@christiaensen2011evolving]. This raises an important question: do rising pollution levels in developing countries threaten agricultural productivity?

@@

Using daily wind data and these spatial variables, I create an identifier separately for every village for whether a given village is located downwind from a coal plant on any given day. I then aggregate this exposure variable based on the desired analysis; for agricultural productivity, for example, I create an exposure variable measuring the total number of exposure days in the five-month agricultural season, focusing only on the monsoon (Kharif) season. More days being located downwind from a high-pollution location leads to substantially higher average concentrations of particulate matter 2.5 (PM 2.5), which is defined as particulate matter with a diameter of less than 2.5 micrometers.

@@

The final set of analyses disentangles the effects of pollution from the effects of coal pollution, specifically, by using wind direction and the rollout of coal plants across the country. Combining instrumental variables with a differences-in-differences strategy, I show that agricultural productivity has an elasticity of around -0.22 with respect to non-coal pollution. The opening of a coal plant, however, doubles the elasticity, to more than -0.44, indicating that coal plant emissions are particularly harmful to agricultural productivity, at least in India. This is consistent with coal emissions being quite different from other forms of pollution; coal emissions contain different types and quantities of harmful substances, meaning effects may differ greatly.

@@

The results are consistent with evidence of effects of pollution on different forms of labor activity. For example, general levels of pollution decrease productivity of call center works in China [@chang2019effect], lead people to perform worse on cognitive functions [@ebenstein2016long; @wen2022lower], lower labor market earnings [@borgschulte2022air], and also lead to lower levels of farm labor productivity in California, driven by changes in ozone concentration [@graff2012impact]. Also in China, @he2019severe find small but significant negative effects of pollution on manufacturing productivity, while @chen2022effect show pollution leads to labor out migration. Relevant to my results, @hanna2015effect show that decreases in pollution lead to increases in labor allocation in Mexico.

### 2023-07-20 draft/draft.tex
updating draft

\title{Air pollution and agricultural productivity in a developing country\footnote{I thank Ariel Ortiz-Bobea, Marc Bellemare, and David Sungho Park for comments and suggestions that greatly improved the quality of the paper, as well as conference attendees at the 2023 KER International Conference. The usual disclaimer applies. All scripts used in this paper are publicly available on my GitHub page: github.com/JoshMerfeld/pollution\_development}}
\author{Joshua D. Merfeld\footnote{KDI School of Public Policy and Management and IZA; \href{mailto:merfeld@kdis.ac.kr}{\nolinkurl{merfeld@kdis.ac.kr}}}}
\date{2023-07-13}

\begin{document}
\maketitle
\begin{abstract}
\noindent I document negative externalities of air pollution in the Indian agricultural sector. Using variation in pollution induced by changes in wind across years, I show that higher levels of pollution lead to decreased agricultural productivity, with large changes in productivity being common. The negative effects of pollution are larger in areas growing more labor-intensive crops, indicating that the pollution works at least partly through direct effects on labor productivity. Finally, combining wind direction with the rollout of coal plants, results indicate that pollution from coal plants has a larger effect on agricultural productivity than other types of pollution. Given that the agricultural sector is a refuge for the poor in many developing countries, these results suggest that the negative externalities of pollution may hit the poorest particularly hard.\\
\strut \\
\textbf{\textit{Keywords}}: pollution, productivity, agriculture, labor, India\\
\textbf{\textit{JEL Codes}}: H40, I15, J22, O13, Q52, Q53
\end{abstract}

@@

While reliance on coal is declining in much of the world, it continues to be an important source of energy in many countries. In 2021, energy generation from coal reached an all-time high,\footnote{www.iea.org/reports/coal-fired-electricity} driven in part by rapid population growth in developing countries, including India. From 1990 to 2010, India's population grew by more than 40 percent, leading to a large increase in demand for power.\footnote{data.worldbank.org} While natural gas has become a popular alternative in much of the world, the Indian government met much of the increased demand through the construction of coal power plants. During those same two decades, power generation from all coal units of at least 30 megawatts in the country more than doubled, from 42.4 gigawatts to more than 100 gigawatts.\footnote{globalenergymonitor.org/}

While there may be positive effects from this increase in power generation -- for example, electricity may lead to economic and productivity growth (Dinkelman 2011; Kline and Moretti 2014; Rud 2012; Van de Walle et al. 2017), though electrification alone may not be a sufficient condition (K. Lee, Miguel, and Wolfram 2020) -- these improvements do not come without downsides, especially when it comes to power derived from coal. Emissions from the production and burning of coal are also directly harmful to human health and the environment; pollutants like sulfur dioxide, nitrogen oxides, methane, and mercury are particularly harmful. While the use of coal had been declining, recent events -- including Russia's invasion of Ukraine -- have led to an increase in the demand for coal.\footnote{www.npr.org/2022/08/15/1117560560/a-rising-demand-for-coal-amidst-war-in-ukraine}

In addition to pollution from coal, pollution more generally is also increasing in much of the world. Carbon dioxide equivalent per capita has almost double over the last 50 years in India, more than doubled in Indonesia, and almost tripled in China.\footnote{\url{https://edgar.jrc.ec.europa.eu}} Much of this increase is due to the development process itself -- as countries develop, their economies transform, with more people moving out of agriculture and into manufacturing and services (Herrendorf, Rogerson, and Valentinyi 2014). Despite this development, agriculture remains a key sector in many developing countries, employing a large proportion of the poor (Dercon 2009). This also means agriculture is a key sector when it comes to poverty reduction (Christiaensen, Demery, and Kuhl 2011). Given the large effects of pollution on health and previously documented impacts on agriculture, this raises a important question: do rising pollution levels in developing countries threaten agricultural productivity?

Although we have made much progress over the last several decades, poverty and inequality remain important challenges in the developing world (Alvaredo and Gasparini 2015). Poverty is particularly prevalent in rural areas, where the majority of the world's poor live (Dercon 2009), and in the agricultural sector, where the majority of the world's poor work (Christiaensen, Demery, and Kuhl 2011). As such, agricultural productivity is a key determinant of poverty in developing countries (Dercon 2009), meaning that any threat to agricultural productivity is also a threat to poverty reduction.

@@

Using daily wind data and these spatial variables, I create an identifier separately for every village for whether a given village is located downwind from a coal plant on any given day. I then aggregate this exposure variable based on the desired analysis; for agricultural productivity, for example, I create an exposure variable measuring the total number of exposure days in the five-month agricultural season, focusing only on the monsoon (Kharif) season. More days being located downwind from a high-pollution location leads to substantially higher average concentrations of particulate matter 2.5 (PM 2.5).\footnote{PM 2.5 is defined as particulate matter with a diameter of less than 2.5 micrometers.}

@@

This paper contributes to several strands of literature. We already have evidence of effects of pollution on different forms of labor activity. For example, general levels of pollution decrease productivity of call center works in China (Chang et al. 2019), lead people to perform worse on cognitive functions (Ebenstein, Lavy, and Roth 2016; Wen and Burke 2022), lower labor market earnings (Borgschulte, Molitor, and Zou 2022), and also lead to lower levels of farm labor productivity in California, driven by changes in ozone concentration (Graff Zivin and Neidell 2012). Also in China, He, Liu, and Salvo (2019) find small but significant negative effects of pollution on manufacturing productivity, while Chen, Oliva, and Zhang (2022) show pollution leads to labor out migration. Relevant to my results, Hanna and Oliva (2015) show that decreases in pollution lead to increases in labor allocation in another developing country.

@@

\leavevmode\vadjust pre{\hypertarget{ref-alvaredo2015recent}{}}%
Alvaredo, Facundo, and Leonardo Gasparini. 2015. {``Recent Trends in Inequality and Poverty in Developing Countries.''} \emph{{Handbook of Income Distribution}} 2: 697--805.

>>>>

\title{Air pollution and agricultural productivity in a developing country\footnote{I thank Ariel Ortiz-Bobea, Marc Bellemare, and David Sungho Park for comments and suggestions that greatly improved the quality of the paper, as well as conference attendees at the 2023 KER International Conference and the University of Reading. The usual disclaimer applies. All scripts used in this paper are publicly available on my GitHub page: github.com/JoshMerfeld/pollution\_development}}
\author{Joshua D. Merfeld\footnote{KDI School of Public Policy and Management and IZA; \href{mailto:merfeld@kdis.ac.kr}{\nolinkurl{merfeld@kdis.ac.kr}}}}
\date{2023-07-19}

\begin{document}
\maketitle
\begin{abstract}
\noindent I document negative externalities of air pollution in the Indian agricultural sector. Using variation in pollution induced by changes in wind across years, I show that higher levels of pollution lead to decreased agricultural productivity. The negative effects of pollution are larger in areas growing more labor-intensive crops, indicating that the pollution works partly through direct effects on labor productivity. I also show that pollution from coal plants has a larger effect on agricultural productivity than other types of pollution. These results suggest that the negative externalities of pollution may hit the poorest particularly hard, given their reliance on agriculture.\\
\strut \\
\textbf{\textit{Keywords}}: pollution, productivity, agriculture, labor, India\\
\textbf{\textit{JEL Codes}}: H40, I15, J22, O13, Q52, Q53
\end{abstract}

@@

In 2021, energy generation from coal reached an all-time high,\footnote{www.iea.org/reports/coal-fired-electricity} driven in part by rapid population growth in developing countries, including India. While natural gas has become a popular alternative in much of the world, the Indian government met much of the increased demand through the construction of coal power plants. From 1990 to 2010, India's population grew by more than 40 percent, leading to a large increase in demand for power.\footnote{data.worldbank.org} During those same two decades, power generation from all coal units of at least 30 megawatts in the country more than doubled, from 42.4 gigawatts to more than 100 gigawatts.\footnote{globalenergymonitor.org/}

While electricity may lead to economic and productivity growth (Dinkelman 2011; Kline and Moretti 2014; Rud 2012; Van de Walle et al. 2017), though electrification alone may not be a sufficient condition (K. Lee, Miguel, and Wolfram 2020), these improvements do not come without downsides, especially when it comes to power derived from coal. Emissions from the production and burning of coal -- pollutants like sulfur dioxide, nitrogen oxides, methane, and mercury, for example -- are also directly harmful to human health and the environment. Although the use of coal had been declining, recent events -- including Russia's invasion of Ukraine -- have led to an increase in the demand for coal.\footnote{www.npr.org/2022/08/15/1117560560/a-rising-demand-for-coal-amidst-war-in-ukraine}

Pollution more generally is also increasing in much of the world. Carbon dioxide equivalent per capita has almost doubled over the last 50 years in India, more than doubled in Indonesia, and almost tripled in China.\footnote{\url{https://edgar.jrc.ec.europa.eu}} Much of this increase is due to the development process itself -- as countries develop, their economies transform, with more people moving out of agriculture and into manufacturing and services (Herrendorf, Rogerson, and Valentinyi 2014). Despite this development, agriculture remains a key sector in many developing countries, employing a large proportion of the poor (Dercon 2009), and agriculture is a key sector for poverty reduction (Christiaensen, Demery, and Kuhl 2011). This raises an important question: do rising pollution levels in developing countries threaten agricultural productivity?

@@

Using daily wind data and these spatial variables, I create an identifier separately for every village for whether a given village is located downwind from a coal plant on any given day. I then aggregate this exposure variable based on the desired analysis; for agricultural productivity, for example, I create an exposure variable measuring the total number of exposure days in the five-month agricultural season, focusing only on the monsoon (Kharif) season. More days being located downwind from a high-pollution location leads to substantially higher average concentrations of particulate matter 2.5 (PM 2.5), which is defined as particulate matter with a diameter of less than 2.5 micrometers.

@@

The results are consistent with evidence of effects of pollution on different forms of labor activity. For example, general levels of pollution decrease productivity of call center works in China (Chang et al. 2019), lead people to perform worse on cognitive functions (Ebenstein, Lavy, and Roth 2016; Wen and Burke 2022), lower labor market earnings (Borgschulte, Molitor, and Zou 2022), and also lead to lower levels of farm labor productivity in California, driven by changes in ozone concentration (Graff Zivin and Neidell 2012). Also in China, He, Liu, and Salvo (2019) find small but significant negative effects of pollution on manufacturing productivity, while Chen, Oliva, and Zhang (2022) show pollution leads to labor out migration. Relevant to my results, Hanna and Oliva (2015) show that decreases in pollution lead to increases in labor allocation in Mexico.

@@
